\newcount\draft\draft=0 % set to 0 for submission or publication
\documentclass[phd,tocprelim]{cornell}
%
% tocprelim option must be included to put the roman numeral pages in the
% table of contents
%
% The cornellheadings option will make headings completely consistent with
% guidelines.
%
% This sample document was originally provided by Blake Jacquot, and
% fixed up by Andrew Myers.
%
%Some possible packages to include
\usepackage{graphics}
\usepackage{moreverb}
%\usepackage{subfigure}
\usepackage{epsfig}
\usepackage{hangcaption}
\usepackage{newtxtext}
\usepackage{newtxmath}
\usepackage{palatino}

% Gator cruft
\let\openbox\relax
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{syntax}
\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{url}
\usepackage{graphicx,pstricks}
\usepackage{xspace}
\usepackage{xxx}
\usepackage{proof}
\usepackage{listings}

\lstset{escapeinside={<@}{@>}} % for color _coding_
\usepackage{booktabs} % for professional tables
\usepackage{subcaption}
\usepackage{bm}

\usetikzlibrary{shapes,arrows, chains}
\usepackage[ligature, inference]{semantic}
\usepackage[T1]{fontenc}
\usepackage{mathpartir}
\usepackage{nccmath}

\newcommand{\mat}{\mathsf{mat}_{n_1{\times}n_2}}
\newcommand{\env}[1]{#1,\sigma}
\newcommand{\defas}{\mathrel{::=}}
\newenvironment{leftalign}%
{\fleqn[5pt]\csname align*\endcsname}%
{\csname endalign*\endcsname\endfleqn}
\newcommand{\alt}{\:|\:}
\newcommand{\bmark}{\textsf}

\mathlig{->}{\rightarrow}
\mathlig{|-}{\vdash}
\mathlig{=>}{\Rightarrow}

% Commutativity Stuff
\usepackage{algorithm2e}
\usepackage{microtype}
\usepackage{fncylab}
\usepackage{float}
\usepackage{balance}
%

\labelformat{algocf}{Algorithm~#1}

\tikzstyle{block} = [rectangle, draw, fill=gray!20,
text width=4em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']

% Code listings configuration.
\lstdefinelanguage{GLSL}{
	keywords={void, float, int, with, auto,
		if, abs, \#ifdef, \#else, \#endif, in, out, uniform,
		varying, attribute, return, void, space, is, coord, canon, type,
		declare, vec, mat, as, has, frame, object, dimension, coord},
	comment=[l]{//},
}
\lstset{
	language=GLSL,
	columns=fullflexible,
	keepspaces=true,
	showspaces=false,
	showstringspaces=false,
	keywordstyle=\bfseries,
	basicstyle=\ttfamily\small,
	basewidth=0.53em,
	aboveskip=0.9\medskipamount,
	belowskip=0.6\medskipamount,
	mathescape=true,
	lineskip={-2pt},
	commentstyle=\color{gray},
	% xleftmargin=\parindent,
}

% Shorthand for inline code.
\newcommand{\code}{\lstinline[keywordstyle=]}

% color definitions
\definecolor{darkolivegreen}{rgb}{0.33, 0.42, 0.18}

% Semantics declarations for the formalism.
\DeclareRobustCommand\tvdash{\mathrel{||}\joinrel\mkern-.5mu\mathrel{-}}
\newcommand{\source}[1]{\llbracket #1 \rrbracket}
\newcommand{\ld}[1]{\leq_\Delta}
\newcommand{\gen}[1]{\langle #1 \rangle} %generic
\newcommand{\sca}[0]{\textrm{scalar}}
\newcommand{\subsumption}{\textrm{SUBSUMPTION}}
\newcommand{\unit}{\textrm{UNIT}}
\newcommand{\scalar}{\textrm{SCALAR}}
\newcommand{\vect}{\textrm{VECTOR}} %both vec and vector lead to funky behavior
\newcommand{\matr}{\textrm{MATRIX}} %can't call this matrix because amsmath
\newcommand{\decl}{\textrm{DECL}}
\newcommand{\assign}{\textrm{ASSIGN}}
\newcommand{\add}{\textrm{ADDITION}}
\newcommand{\smul}{\textrm{SCAL MUL}}
\newcommand{\vmul}{\textrm{VEC MUL}}
\newcommand{\mmul}{\textrm{MAT MUL}}
\newcommand{\vcmul}{\textrm{VEC CMUL}}
\newcommand{\mcmul}{\textrm{MAT CMUL}}

\newcommand{\IH}{\textrm{I.H.}}
\newcommand{\TC}{\textrm{T.C.}}
\newcommand{\sub}{\textrm{SUBST}}

\newif\ifsemanticsdoc

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

%if you're having problems with overfull boxes, you may need to increase
%the tolerance to 9999
\tolerance=9999

\bibliographystyle{plain}
%\bibliographystyle{IEEEbib}

\renewcommand{\caption}[1]{\singlespacing\hangcaption{#1}\normalspacing}
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.75}

\title {Language Designs for Geometry and Heterogeneous Reasoning in Graphics Programming}
\author {Dietrich Geisler}
\conferraldate {August}{2024}
\degreefield {Ph.D.}
\copyrightholder{Dietrich Geisler}
\copyrightyear{2024}

\begin{document}

\maketitle
\makecopyright

% Enable our \mathlig substitutions.
\mathligson

\begin{abstract}
There has been growing demand for graphical image rendering in the past several decades.  This demand has arisen primarily in video games, but also from fields as broad as film, art, architecture, and scientific simulation.  A major challenge with expanding use of rendering, however, is that graphics programming is difficult, requiring significant field expertise when abstractions break down.

In this dissertation, we will examine how we may be able to design programming languages to ameliorate some of these challenges.  Our goal will be to examine two specific difficulties in graphics programming reasoning: geometric correctness and performance in heterogeneous device communication.  

In the first chunk of this dissertation, we will examine Gator, a language which provides type-level reasoning for a class of bugs we describe as ``geometry bugs", as well as a lightweight mechanism to reason about operations on geometry.  In the second chunk of this dissertation, we will discuss Caiman, a language which typechecks heterogeneous implementations against a fixed specification.  We will also examine how Caiman's type-level restrictions allow for separating performance and correctness, as well as providing a mechanism for restricted synthesis of heterogeneous programs.
\end{abstract}

\begin{biosketch}
Your biosketch goes here. Make sure it sits inside
the brackets.
\end{biosketch}

\begin{dedication}
For my siblings: Conrad, Emil, and Claire
\end{dedication}

\begin{acknowledgements}
This PhD would not have been possible without the support of mentors, colleagues, friends, and family:
First and foremost, I would like to thank my advisor, Adrian Sampson.  Without his mentorship, encouragement, and support, I would not be here today.  Beyond research, Adrian has helped me find life direction, and how much his guidance means to me cannot be overstated.

The second chapter of this dissertation was written in collaboration with Irene Yoon, Aditi Kabra, Horace He, and Yinnon Sanders.  Irene in particular provided a huge amount of work and produced some excellent design for building Linguine, the excellently--named predecessor to the Gator language.  

Additionally, Evan Adler, Kimberly Baum, Ben Gillott, Henry Liu, and Palini Ramnarayan all provided important contributions to Gator systems and tooling, and worked on excellent projects that helped refine our understanding of Gator usability.

The third chapter of this dissertation was written in direct collaboration with Aditi Kabra, who also detailed much of the theory of Gator.

While I do not have specific writing to credit, thank you to Haoxuan Chen, Meredith Hu, Paul Joo, and David Siher for their discussions and contributions to this body of research.  Their work was invaluable for helping refine our research direction and explore topics I wish we could have expanded further.

The fourth chapter of this dissertation, Caiman, could only have been written in collaboration with Oliver Daids, who both completely changed how we approached the problem presented by Caiman \textit{and} worked out details for the complex theory and interlocking systems needed to realize the Caiman compiler.  Stephen Verderame provided invaluable contributions in expanding Caiman with the frontend language presented in the paper, and the explanations provided here could not be given without his hard work.

Caiman could also not be what it is today without the work provided by Mia Daniels, Mateo Guynn, and Patrick LaFontaine, all of whom contributed systems engineering and design to the sprawling Gator project.

I would like to thank Chris Batten, Nate Foster, Steve Marschner, Jos\'e Mart\'inez, and Walker White for their mentorship and support in producing this research and giving me both motivation and direction.  This dissertation would not be the same without such diverse ideas and discussions.

I could not have finished this work without the support of my close friends: Gemma Clark, Jarem Kilby, Elliot Lee, Haobin Ni, Oliver Richardson, and Michael Roberts.  I keep this paragraph short lest it take up the rest of what was supposed to be a dissertation, but I could not have done this work without your support.  Thank you.

Thank you to my siblings, Conrad, Emil, and Claire for your love and support.  Thank you to my Dad, who has stayed a fixture of my life despite geographic distance.

Finally, I must end by thanking my Mom, who has somehow raised and supported both me and my two brothers as a single parent for all these years.  I truly could not have done this work without you.

\end{acknowledgements}

\contentspage
\tablelistpage
\figurelistpage

\normalspacing \setcounter{page}{1} \pagenumbering{arabic}
\pagestyle{cornell} \addtolength{\parskip}{0.5\baselineskip}

\chapter{Introduction}

\include{introduction}

\chapter{Gator: Geometry Types for Graphics Programming}

\input{gatormain}

\chapter{Online Verification of Commutativity}

\input{commmain}

\chapter{Caiman: DSL for Optimizing Heterogeneous Program Communication (Caiman)}

\input{caimanmain}

\chapter{Conclusion and Future Directions}

We have examined how domain--specific language design for graphics programming can help programmers with managing correctness and performance considerations.  With Gator, we have shown how we can define a specialized type system for reasoning about geometric types, and how these type definitions can provide information for multi--operation transformations.  Additionally, we have shown how Gator exposed a surprising challenge in reasoning about equivalence of these transformations through the work on commutative diagrams.

With the Caiman language, on the other hand, we examine how definitions of correctness and equality with respect to a specification can help explore performant solutions.  Additionally, we have shown how we can provide decomposable program implementations for heterogeneous code from these fixed specifications through explication.

Perhaps the most valuable work that should be done is validating the use of these approaches in a more practical setting.  While we have provided evidence of usability for Gator and Caiman, it can be difficult to rely on intuition for the practicalities of these designs, particularly in larger--scale industrial projects.  How do users interact with Gator types and Caiman language features?  Can beginners realistically learn and apply these systems?  How well does Caiman's approach scale once compilation speed is a realistic concern?  These are all questions that may require an academic user study or larger-scale tooling.

To emphasize this second direction more, both Gator and Caiman would have significantly more impact and potential use with better tooling and integration into existing applications.  Rewriting a project with hundreds of thousands of lines of code in either of these settings is clearly unrealistic, and while Gator and Caiman both are designed to ``slot in" to an existing codebase, both type systems are much more useful when given more information about the structure of a system.

Concretely, this sort of effort simply requires more engineering.  There are, however, several interesting research and design questions within this engineering.  Could we meaningfully build a (partial) C-to-Caiman compiler to help with optimizing existing programs, while preserving the Caiman semantic representation?  Could we partially annotate a large--scale graphics program with Gator--style geometry types and have inference be ``good enough" to be useful?  Caiman's transparency seems to lend itself to some interesting usability cases -- could we build a specialized extension for Caiman to ``view" the code that the explicator produces?

\appendix

\chapter{Gator Appendix}
\input{gator-appendix}

\chapter{Caiman Appendix}
\input{caiman-appendix}

\bibliography{refs}

\end{document}
